---
title: "Regression and Beers"
author: "Wells"
date: "2025-12-02"
output: html_document
---

This is another exercise to break down how a simple linear regression works and how to use it in real life.
The topic: beer.

This following exercise comes from Diez, D., Cetinkaya-Rundel, M., & Barr, C. (2024). OpenIntro Statistics (4th ed.). http://www.openintro.org/redirect.php?go=os&referrer=os4_pdf, p 335


Many people believe that gender, weight, drinking habits, and many other factors are much more important in predicting blood alcohol content (BAC) than simply considering the number of drinks a person consumed. Here we examine data from sixteen student volunteers at Ohio
State University who each drank a randomly assigned number of cans of beer. These students were evenly divided between men and women, and they differed in weight and drinking habits. Thirty minutes later, a police officer measured their blood alcohol content (BAC) in grams of alcohol per deciliter of blood.



Load libraries
```{r}
library(tidyverse)
```


Load data
```{r}
bac_data <- read_csv("https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/activities/bac%20regression/bac_data.csv")

bac_data

#Student #1 drank 5 beers and has a blood alcohol level of 0.13, etc
```


# Linear Regression
```{r}
# Run the regression
model <- lm(bac ~ beers, data = bac_data)

# Display regression results
cat("\n\nRegression Results:\n")
print(summary(model))
```

 Plot the data and the model line
```{r}
# Create regression equation text
eq_text <- sprintf("BAC = %.4f + %.4f * beers", 
                   coef(model)[1], coef(model)[2])


ggplot(bac_data, aes(x = beers, y = bac)) +
  geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1) +
  labs(
    x = "Number of Beers",
    y = "Blood Alcohol Content (BAC)",
    title = "BAC vs Number of Beers Consumed"
  ) +
  annotate("text", x = min(bac_data$beers), y = max(bac_data$bac), 
           label = eq_text, hjust = 0, vjust = 1, size = 4) +
  theme_minimal()

```

**Task 1: Interpret the results of this regression. Answer the following questions:**

a) What the predictor or independent variable? What is the response or dependent variable? 
b) What is the regression model asking and what does it tell us?
c) Do the data provide strong evidence that drinking more cans of beer is associated with an increase in blood alcohol? 
d) State the null and alternative hypotheses, report the p-value, and state your conclusion.


**Answer:**
a) The independent variable is the number of beers, and the response is the Blood Alcohol Content.
b) The regression model is asking how the BAC (the response) changes as the number of beers (the independent variable) changes, and what might that look like through a straight-lined relationship.
c) The data provides strong evidence that drinking more cans of beer is associated with an increase in blood alcohol. This is clear in the results, which displays three asterisks because the p-value is very low, meaning high significance. This implies that there is strong evidence showing that the number of beers and Blood Alcohol Content are correlated. 
d) 
  Null: The number of beers consumed and Blood Alcohol Content are not significantly correlated, meaning that for each can of beer, BAC would increase by less than 0.001 points.
  Alternative: The number of beers consumed and Blood Alcohol Content are significantly correlated, meaning that for each can of bear, BAC would increase by more than 0.001 points.
  p-value: 0.0005239
  Conclusion: There is strong evidence showing that the number of beers and Blood Alcohol Content are correlated, meaning that for each can of beer, the BAC would increase by over 0.001 points.



