---
title: "Diversity Index for Maryland Cities"
date: "Nov 8, 2025"
author: "Mara K. Bell"
---

**Mara K. Bell**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

In this exercise, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

A score of 0 indicates no diversity (everyone is the same race/ethnicity), while a score closer to 100 indicates higher diversity. The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

## Task 1: Load and Describe the Data (3 points)

Run the following code to load the data, then examine it and describe what it contains. Based on the raw numbers alone, which cities stand out as diverse to you?



```{r load-data}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/usat_diversity_index.csv")

# Preview structure
View(md_cities)
```

**
- This dataset contains several columns, including the total population of each city and the amount of people in each city by race.
- Based on the raw numbers alone, cities like Baltimore City, Germantown, and Columbia all stand out as diverse to me. This seemed likely after sorting the data set to display the cities with the most people with two or more races, which I think is a good indicator of a diverse population. Baltimore City had the most at 14,123 people. Germantown had the second most people with two or more races at 4,235 people, and Columbia had 3,961 people of two or more races. 
- At the same time, we should also consider the diversity rate or race percentage per population as some cities have bigger populations than others. 
**


## Task 2: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities |>
  mutate(
    White_pct = White / TotalPop,
    Black_pct = Black / TotalPop,
    Native_pct = Native / TotalPop,
    Asian_pct = Asian / TotalPop,
    NHPI_pct = NHPI / TotalPop
  )
```

## Task 3: Square the Race Percentages

We need to square the race percentages in order to calculate the probability that any two people chosen at random from a given census area are of different races or are Hispanic or not.

```{r square-race}
md_cities <- md_cities |>
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    Native_pct_sq = Native_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```


## Task 4: Calculate Race Probabilities (3 points)

The race probability is the sum the squared probabilities for each of the races. It represents the probability that two people randomly chosen from the same city will be of the same race. Which city has the highest race probability, and why do you think that is?


```{r race-prob}
md_cities <- md_cities |>
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + Native_pct_sq + Asian_pct_sq + NHPI_pct_sq
  ) |> 
  arrange(desc(Race_Prob))

View(md_cities)
```

**
- It seems that Langley Park (0.04), Wheaton (0.18), and Glenmont (0.19) have the lowest race probabilities, and that Summerfield (0.87), Lake Shore (0.87), and Hillcrest Heights (0.87) have the highest race probabilities.
- Summerfield is the city with the heighest race probability at 0.8715 meaning that if two people were chosen randomly from the same city, they would likely be of the same race. 
- This means that the city is the least diverse because of the predominantly Black population in the city and very low race percentages in all categories. 
- This causes the race probability to be very high, given that if two people were chosen, they'd likely be Black most of the time.
**

## Task 5: Calculate Ethnicity Percentages and Their Squares (3 points)

Hispanic origin is not a race but an ethnicity or nationality – and a separate Census question. The probability someone is Hispanic or not Hispanic must be calculated separately from race, and Hisp_Prob is the probability that two randomly selected people will be Hispanic or not. That's an important distinction, because unlike race probability, this measures the likelihood of a binary question: that both random people will be Hispanic. What do you notice about the cities with the *lowest* Hisp_Prob scores?


```{r ethnicity}
md_cities <- md_cities |>
  mutate(
    HispTotal = HISPANIC + NotHispanic,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NotHispanic / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  ) |> 
  arrange(Hisp_Prob)

View(md_cities)
```

**Cities with the lowest Hisp_Prob scores, like Chillum, East Riverdale, and Wheaton had race percentages of a race that was not Hispanic that outweighed the Hispanic populations, therefore making it less probable that both random people chosen would be Hispanic. For example, with Chillum, the predominant race was Black at 18,324 people while the Hispanic population had 16,474 people, which was less.**

## Task 6: Multiply Race and Ethnicity Probabilities

Here we multiply the two probabilities together to get the probability that any two random people have the SAME race and ethnicity.

```{r combined-prob}
md_cities <- md_cities |>
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```


## Task 7: Calculate Diversity Index (3 points)

The final step is to calculate the probability that two people randomly chosen WILL NOT BE OF THE SAME RACE OR ETHNICITY. In other words, the highest index scores will be cities where those two random people will be from different backgrounds. Put simply, more diverse. Which cities are the top 3, and where are they located? Look at the bottom of the list - which city is the least diverse, and where is it?


```{r diversity-index}
md_cities <- md_cities |>
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  ) |> 
  arrange(desc(Diversity_Index))

View(md_cities)
```

**
- The top 3 cities with highest diversity index are Langley Park at 97.1, Wheaton at 90.1, and Adelphi at 90. Langley Park and Adelphi are in Prince George's County and Wheaton is in Montgomery County, neighboring counties in Maryland. The least diverse cities are Lake Shore at 15.5, Lake Arbor at 16.7, and Largo at 17.8. Lake Arbor and Largo are in Prince George's County and Lake Shore is in Anne Arundel County. 
- This is very interesting that Prince George's County cities show up for both the most and least diverse cities and Maryland.
**

### Pitch a story (8 points)

Given the data that you've produced here, write a headline and lede of the most interesting story from this exercise, and add below that details on how you would report it out: who you'd talk to and the questions you'd ask. When you describe the Diversity Index values, do so in a similar way to how USA Today did; the raw values themselves won't mean much to readers.

**
- Headline: Langley Park is Maryland's #1 most diverse city, study shows. Was it always this way?
- Lede: Langley Park ranks #1 as the most diverse city, with Wheaton and Adelphi trailing behind, according to a USA Today study by factors like racial diversity and population sizes.
- Reporting: I would love to make this a feature story and talk to current residents now and how they feel about living in a city that's ranked the most diverse in the state. Then, I would go to older residents who have lived in the area for longer amounts of time and ask them if they've seen changes like growth of diversity. I would also look at previous census data to see how much diversity has grown across a decade or a couple of decades. 
**
