---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

**A:**
Our group's mean was not close at all. The brown M&Ms took up 11% of the total batch of M&Ms, the red M&Ms took up 14%, and the rest of the colors took up 75%.

Different groups get different results because each sample was random and only a portion of the bag (a smaller sample size).

When we combine all groups individual extreme results no longer show as much and are instead, normalized. This is because the sample size would be larger, therefore reducing sampling error and providing a more accurate view of the data set.

This relates to polls and surveys in real life as pollsters aim to collect data from large, random samples because they provide more reliable results than a small sample, which is more likely to have sampling errors.

------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:**
Only 23% of UMD community members responded to the survey, or 12,233 people, a drastic difference between the population and the sample. Undergraduate students were underrepresented in this sample as only 20% of students responded to the survey, compared to the population where undergraduate students take up a little over half of the UMD community. Data collectors aimed to identify the differences between those who took the survey through demographic variables, including racial/ethnic identity and gender. A way the survey attempted to deal with differences between the population and sample is through combining racial and ethnic groups with not enough respondents into groups, like "Races/Ethnicities with Small Sample Sizes." I find that the document did not go in depth enough about how the survey attempted to deal with the differences in racial and ethnic identities displayed in the sample compared to the population. How did the survey decide on what racial/ethnic groups should be included? Did they base their choices off of the U.S. Census?

------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(42)
adults <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(adults)
```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <- adults |>
  sample_n(100)
```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
adults |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```


### Task 6: Compare the sample and population statistics
**A**

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

**A**: 
The means and standard deviations are relatively similar between the sample and the population, meaning that the random sampling allowed for a more accurate representation of the data. If I run the calculations with a smaller sample size, as below, the mean, median and standard deviation is a lot less accurate to the population as it would be with a larger sample size.

```{r}
sample_35 <- adults |>
  sample_n(35)

sample_35 |> summarize(mean = mean(age), median = median(age), sd = sd(age))
```

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
