---
title: "UMD Fees"
date: 8-25-2025
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse, janitor
Janitor performs important data cleaning operations

```{r}
library(tidyverse)
#install.packages('janitor')
library(janitor)
```




### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv") |> 
   clean_names()
```

### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (fall_2024- fall_2021)/fall_2021 * 100)
```


### Task 4: Calculate the mean and median fees for Fall 2024 using `mutate()`

```{r}

umd_fees |> 
    summarize(mean_2024_fees = mean(fall_2024),
            median_2024_fees = median(fall_2024)
)


```

### Task 5: Add to the previous the max and minimum values for 2024 in addition to the mean and median fees for Fall 2024 using `mutate()`

```{r}

umd_fees |> 
    summarize(mean_2024_fees = mean(fall_2024),
            median_2024_fees = median(fall_2024),
            max_2024_fees = max(fall_2024),
            min_2024_fees = min(fall_2024)
)


```

### Task 6: Describe our results

Write a few sentences about what you found in the data. What's the story here?

# In the data, we found that the Student Sustainability sector, there was the largest percent change across the years at 150%. It's worth reporting on that change and seeing the growth (or lackthereof) in the department and its' sustainability goals and projects, along with fleshing out where the money went. Some things also worth noting are the athletics department's flat fees and how else the department gets and allocates their money. Recreation Services also has the highest fees at the university.

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
