---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv") |> as.data.frame()
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

ANSWER: Some of the categories of incidents in this data set include robberies, sexual offenses, automobile thefts, breaking and entering, assault, and homicide. The data is specified in many incidental categories. For example, their is a distinction between a commercial robbery versus a residential robbery. From a reporting standpoint, some categories like sexual offenses and more violent crimes like homicides are more important to communities (based on headlines and readership) and therefore, reporters would likely pay attention to those categories. These categories, in my eyes, would actually narrow the data for a reporter, therefore making it less difficult to summarize the data.

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> 
  mutate(total_incidents =  rowSums(across(2:13))) |> 
  mutate(average_incidents = total_incidents/12)

```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

ANSWER: After examining the new dataframe, accidents, stolen vehicles and theft had the highest average number of incidents per month. Some of the lowest were residential robberies, homicide and sexual offenses. This was interesting to me because I named two of the lowest occurring incidents in PG County as the most likely to be newsworthy earlier. This data shows a pattern of vehicle-related crimes, a statistic that measures up with most crimes across the country, which are often related to vehicles. I'm curious to see if the high incidents in stolen vehicles and thefts from vehicles is common among neighborhing counties. If it isn't, I believe this would be a story to look into. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> 
  mutate(june_change =  (Jun - average_incidents)/average_incidents * 100) |> 
  mutate(july_change =  (Jul - average_incidents)/average_incidents * 100) |>
  mutate(august_change =  (Aug - average_incidents)/average_incidents * 100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

ANSWER: I do not see a clear pattern while looking at this dataframe. Some of the percentage change calculations are all over the place, with some ranging as low as -45% to as high as 65%. I don't believe that this calculation is a representation of the theory that crime is more likely to occur in the summer months. 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER:

Vehicle-related crimes remain Prince George's County's most prevalent incidents, according to crime data from the county's police department. 

According to 2024 crime data released by the department, stolen vehicles and theft from vehicles top the charts when surveying crime in the area, a widespread issue in the county, averaging at nearly 400 incidents each throughout the year. The statistic comes after a former police officer in the county was charged for driving stolen vehicles in October 2024, with charges later dropped in May.


### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
