---
title: "HW3 WH Salaries"
Mara K. Bell
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv") |>
  clean_names()
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER: Looking at the data, columns include the name of a White House staffer, their status, their salary, and their title. There are only two statuses, "Employee" or "Detailee" (which the latter means an employee temporarily assigned to another position). The salaries range from zero to as high as around 250,000 dollars. From a reporting standpoint, I'd like to look at why people of the same position are paid drastically differently, for example, for the position "Associate Counsel." Salary values of zero could cause problems for analysis as they are extremely low values. Michelle Barran's salary could also cause issues, being the highest by a wide difference at 251258 dollars.


### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> 
    summarize(mean_salary = mean(salary),
            median_salary = median(salary)
)
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> 
    summarize(mean_salary = mean(salary),
            median_salary = median(salary),
            min_salary = min(salary),
            max_salary = max(salary)
)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER: Based on my analysis, the mean and median salary being nearly the same means that the distribution of the data is nearly symmetrical instead of skewed left or right. This means that there are likely no outliers in the data that would affect the mean. The mean and median salary also  explain that data points at the very low or very high point in the dataset are relatively under or overpaid compared to the average salary. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER: Based on the data, I believe my best story idea is to examine the salaries and people in positions with greatly differing salaries. For example, the position "Senior Policy Advisor," with some paid at 88,000 dollars while others are paid at nearly 200,000 dollars. The same is seen for the position, "Associate Counsel" where some are paid at a 70,000 dollar difference compared to others in the same position. My next step would be to do basic research, then reach out to press officers for Human Resources and specific staffers to learn about feelings toward salaries. 
